\textbf{Evaluating RISC-V Enclaves:}\\ Security Without the Slowdown?

### 2025-09-04 draft_presentation.tex
added overlays

% Keystone 
\begin{frame}{From Problem to Solution: Enter Keystone}
\begin{itemize}
    \item Modular, open-source TEE for RISC-V.
    \item Uses Physical Memory Protection (PMP) and minimal Security Monitor for isolation.
    \item Low Trusted Computing Base (TCB) for security and flexibility.
    \item Optimized for secure computation in untrusted environments.
\end{itemize}
\begin{block}{Key Concept}
\textit{Low TCB reduces attack surfaces while enabling high flexibility and performance in enclave-based systems.}
\end{block}
\end{frame}

% Keystone Framework Architecture
\begin{frame}{Keystone Framework Architecture}
\begin{block}{Overview}
    Keystone is designed to be modular, with clear isolation between enclaves and the OS.
\end{block}

%\resizebox{\linewidth}{!}{%
\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{/home/individual-01/Documents/keystone-thesis/figures/keystone_overview.png}
%\caption{Keystone architecture showing components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
\label{fig:keystone_grouped}
\end{figure}
\end{frame}

% Keystone vs Other TEEs
\begin{frame}{Keystone vs. Other TEEs}
\begin{figure}[htbp]
\centering
\renewcommand{\arraystretch}{1.3} % more row height
\setlength{\tabcolsep}{5pt}       % more column padding
\rowcolors{2}{gray!5}{white}      % alternating row colors

\begin{tabular}{@{}lcccc@{}}
\toprule
\rowcolor{gray!20}
\textbf{Feature} & \textbf{Keystone} & \textbf{SGX} & \textbf{TrustZone} & \textbf{SEV} \\
\midrule
Open Source            & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{red!15}No & \cellcolor{red!15}No \\
Dynamic Enclaves       & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{yellow!15}Limited & \cellcolor{red!15}No \\
Hardware Dependency    & \cellcolor{green!15}RISC-V & \cellcolor{red!15}Intel & \cellcolor{red!15}ARM & \cellcolor{red!15}AMD \\
TCB Size               & \cellcolor{green!15}Low & \cellcolor{yellow!15}Medium & \cellcolor{red!15}High & \cellcolor{yellow!15}Medium \\
Flexible Configurations& \cellcolor{green!15}High & \cellcolor{red!15}Low & \cellcolor{red!15}Low & \cellcolor{yellow!15}Medium \\
\bottomrule
\end{tabular}
\caption{Feature comparison of major TEEs.}
\label{fig:tee-features}
\end{figure}

\end{frame}

@@

\begin{frame}{Enclave Overhead and System Behavior}
\small
\begin{block}{Performance Overhead}
\begin{itemize}
    \item \textbf{Enclave execution} introduces overhead: 10\% to 15\%.
    \item \textbf{Compute-bound workloads} (CoreMark, Dhrystone) see minimal performance loss.
    \item \textbf{Memory-bound workloads} (Kyber) experience severe slowdowns due to added security checks and memory isolation.
\end{itemize}
\end{block}

\begin{block}{Overhead Breakdown}
\begin{itemize}
    \item Memory isolation via PMP adds fixed latency per memory access.
    \item Context switches between host and enclave domains are costly.
\end{itemize}
\end{block}

\centering
\textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
\end{frame}

\begin{frame}{Performance Trends and Hardware Bottlenecks}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=14pt,
    width=10cm,
    height=6cm,
    ymin=0, ymax=120,
    ylabel={Relative performance (\%)},
    symbolic x coords={CoreMark,Dhrystone,Kyber},
    xtick=data,
    ymajorgrids,
    grid style=dashed,
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1}
]

@@

\legend{Enclave,Native}
\end{axis}
\end{tikzpicture}
\caption{Relative performance of enclave vs native execution.}
\end{figure}
\end{frame}

@@

\begin{frame}{PMP Entry Exhaustion and Multi-Enclave Scalability}
\small
\begin{block}{PMP Resource Limitations}
\begin{itemize}
    \item \textbf{Maximum PMP entries}: 16.
    \item Exhausted PMP entries \textbf{prevent new enclave creation}, causing errors even with abundant system resources.
    \item Failure due to PMP exhaustion: \texttt{SBI error codes}.
\end{itemize}
\end{block}

\begin{block}{Scalability Bottleneck}
\begin{itemize}
    \item In \textbf{multi-tenant environments}, PMP exhaustion limits the number of concurrent enclaves.
    \item This creates scalability issues, especially for high-concurrency workloads.
\end{itemize}
\end{block}

\centering
\textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
\end{frame}

@@

\begin{frame}{PMP Entries vs Max Concurrent Enclaves}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=10cm,
    height=6cm,
    xlabel={PMP Entries},
    ylabel={Concurrent Enclaves},
    xmin=0, xmax=16,
    ymin=0, ymax=16,
    xtick={0,4,8,12,16},
    ymajorgrids,
    grid style=dashed,
    mark options={solid},
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    clip=false,
    cycle list={{teal,mark=*,thick}},
    tick label style={font=\small},
    label style={font=\small\bfseries},
]

@@

\end{axis}
\end{tikzpicture}
\caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
\end{figure}
\end{frame}

\begin{frame}{PMP Misalignment: Why It Matters}
\begin{block}{The Problem}
PMP entries must describe power-of-two, naturally aligned regions (NAPOT). Misaligned memory allocations cause unnecessary padding.
\end{block}

\begin{block}{The Effect}
\begin{itemize}
    \item Misalignment causes a single memory region to consume multiple PMP entries.
    \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
\end{itemize}
\end{block}

\vspace{1em}
\centering
\textbf{Next: Visualizing the impact →}
\end{frame}

\begin{frame}{PMP Misalignment: Visualized}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]

@@

\vspace{0.4em}
\scriptsize
\textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
\end{figure}
\end{frame}

@@

\begin{frame}{Deployment Takeaways}
\begin{block}{Best Practices}
\begin{itemize}
    \item Use NAPOT-aligned, contiguous memory regions.
    \item Consolidate workloads into fewer enclaves.
    \item Batch calls to reduce context switches.
\end{itemize}
\end{block}

\begin{alertblock}{Key Principle}
Balance performance, isolation, and PMP resource use.
\end{alertblock}
\end{frame}

@@

\begin{frame}{Workload-Aware Enclave Design}
\begin{block}{Hybrid Cryptographic Workloads}
\begin{itemize}
    \item Keep keys inside enclaves.
    \item Offload bulk crypto outside to reduce overhead.
\end{itemize}
\end{block}

\begin{block}{Target Use Cases}
\begin{itemize}
    \item Isolate sensitive logic (e.g., ML inference, attestation).
    \item Avoid enclaves for memory-bound workloads like Kyber.
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{Enclave Concurrency and Multiplexing}
\begin{block}{Problem}
Too many active enclaves → PMP exhaustion, context switch cost.
\end{block}
\begin{alertblock}{Solution}
Use enclave multiplexing: combine tasks into fewer enclaves.
\end{alertblock}
\begin{block}{Bonus}
Dynamic PMP reallocation can further improve scalability.
\end{block}
\end{frame}

@@

% Future Work and Improvements
\begin{frame}{Hardware Outlook \& Future Directions}
\begin{block}{Limitations}
\begin{itemize}
    \item Few PMP entries.
    \item Rigid NAPOT region encoding.
\end{itemize}
\end{block}

\begin{alertblock}{Future Opportunities}
\begin{itemize}
    \item More flexible PMP schemes.
    \item Crypto accelerators for enclave workloads.
    \item Software-defined memory isolation models.
\end{itemize}
\end{alertblock}
\end{frame}

>>>>

% Keystone 
\begin{frame}{From Problem to Solution: Enter Keystone}
    \only<1>{
\begin{block}{What is Keystone?}
\begin{itemize}
    \item Modular, open-source TEE for RISC-V.
    \item Uses Physical Memory Protection (PMP) and minimal Security Monitor for isolation.
    \item Low Trusted Computing Base (TCB) for security and flexibility.
    %\item Optimized for secure computation in untrusted environments.
\end{itemize}
\end{block}
    }
% \begin{block}{Key Concept}
% \textit{Low TCB reduces attack surfaces while enabling high flexibility and performance in enclave-based systems.}
% \end{block}
%\end{frame}

% Keystone Framework Architecture
%\begin{frame}{Keystone Framework Architecture}
%     \only<1>{
% \begin{block}{Overview}
%     Keystone is designed to be modular, with clear isolation between enclaves and the OS.
% \end{block}
%     }
%     \only<2>{
%\resizebox{\linewidth}{!}{%
\only<2>{
\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{/home/individual-01/Documents/keystone-thesis/figures/keystone_overview.png}
\caption{Keystone architecture showing components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
\label{fig:keystone_grouped}
\end{figure}
    }
%\end{frame}

% Keystone vs Other TEEs
%\begin{frame}{Keystone vs. Other TEEs}
\only<3>{
\begin{figure}[htbp]
\centering
\renewcommand{\arraystretch}{1.3} % more row height
\setlength{\tabcolsep}{5pt}       % more column padding
\rowcolors{2}{gray!5}{white}      % alternating row colors

\begin{tabular}{@{}lcccc@{}}
\toprule
\rowcolor{gray!20}
\textbf{Feature} & \textbf{Keystone} & \textbf{SGX} & \textbf{TrustZone} & \textbf{SEV} \\
\midrule
Open Source            & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{red!15}No & \cellcolor{red!15}No \\
Dynamic Enclaves       & \cellcolor{green!15}Yes & \cellcolor{red!15}No & \cellcolor{yellow!15}Limited & \cellcolor{red!15}No \\
Hardware Dependency    & \cellcolor{green!15}RISC-V & \cellcolor{red!15}Intel & \cellcolor{red!15}ARM & \cellcolor{red!15}AMD \\
TCB Size               & \cellcolor{green!15}Low & \cellcolor{yellow!15}Medium & \cellcolor{red!15}High & \cellcolor{yellow!15}Medium \\
Flexible Configurations& \cellcolor{green!15}High & \cellcolor{red!15}Low & \cellcolor{red!15}Low & \cellcolor{yellow!15}Medium \\
\bottomrule
\end{tabular}
\caption{Feature comparison of major TEEs.}
\label{fig:tee-features}
\end{figure}
}
\end{frame}

@@

\begin{frame}{Enclave Overhead and System Behavior}
\small
% First overlay: text only
\only<1>{
\begin{block}{Performance Overhead}
\begin{itemize}
    \item \textbf{Enclave execution} introduces overhead:
        \begin{itemize}
            \item Compute-bound workloads like CoreMark and Dhrystone incur moderate overhead of 1.06$\times$ to 1.61$\times$.
            \item Hybrid workloads (Compute- and memory-bound) like Kyber suffer severe overhead, up to 2.73$\times$.
        \end{itemize}
    %\item \textbf{Memory-bound workloads} experience significant penalties due to enclave-induced memory isolation and security checks.
\end{itemize}
\end{block}
}
% Second overlay: add figure
\only<2>{
\begin{block}{Overhead Breakdown}
\begin{itemize}
    \item Memory isolation via PMP adds fixed latency per memory access.
    \item Context switches between host and enclave domains are costly.
\end{itemize}
\end{block}
}
% \centering
% \textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
% \end{frame}
% Figure appears only on the second slide of this frame
\only<3>{
% \begin{frame}{Performance Trends and Hardware Bottlenecks}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=14pt,
    width=10cm,
    height=6cm,
    ymin=0, ymax=120,
    ylabel={Relative performance (\%)},
    symbolic x coords={CoreMark,Dhrystone,Kyber},
    xtick=data,
    ymajorgrids,
    grid style=dashed,
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1}
]

@@

\legend{Native,Enclave}
\end{axis}
\end{tikzpicture}
\caption{Relative performance of enclave vs native execution.}
\end{figure}
}
\end{frame}

@@

\begin{frame}{PMP Entry Exhaustion and Multi-Enclave Scalability}
\small
\only<1>{
\begin{block}{PMP Resource Limitations}
\begin{itemize}
    \item \textbf{Maximum PMP entries}: 16.
    \item Exhausted PMP entries \textbf{prevent new enclave creation}, causing errors even with abundant system resources.
    \item Failure due to PMP exhaustion: \texttt{SBI error codes}.
\end{itemize}
\end{block}
}
\only<2>{
\begin{block}{Scalability Bottleneck}
\begin{itemize}
    \item In \textbf{multi-tenant environments}, PMP exhaustion limits the number of concurrent enclaves.
    \item This creates scalability issues, especially for high-concurrency workloads.
\end{itemize}
\end{block}
}

% \centering
% \textbf{Key Insight:} PMP limitations represent a significant scalability bottleneck for RISC-V enclave-based systems.
% \end{frame}

@@

% \begin{frame}{PMP Entries vs Max Concurrent Enclaves}
% \centering
\only<3>{
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=10cm,
    height=6cm,
    xlabel={PMP Entries},
    ylabel={Concurrent Enclaves},
    xmin=0, xmax=16,
    ymin=0, ymax=16,
    xtick={0,4,8,12,16},
    ymajorgrids,
    grid style=dashed,
    mark options={solid},
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    clip=false,
    cycle list={{teal,mark=*,thick}},
    tick label style={font=\small},
    label style={font=\small\bfseries},
]

@@

\end{axis}
\end{tikzpicture}
\caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
\end{figure}
}
\end{frame}

\begin{frame}{PMP Misalignment: Why It Matters}
    \only<1>{
\begin{block}{The Problem}
PMP entries must describe power-of-two, naturally aligned regions (NAPOT). Misaligned memory allocations cause unnecessary padding.
\end{block}
    }
%     \only<2>{
% \begin{block}{The Effect}
% \begin{itemize}
%     \item Misalignment causes a single memory region to consume multiple PMP entries.
%     \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
% \end{itemize}
% \end{block}

% \vspace{1em}
% \centering
% \textbf{Next: Visualizing the impact →}
%\end{frame}
%     }
    \only<2>{
%\begin{frame}{PMP Misalignment: Visualized}
%\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]

@@

\vspace{0.4em}
\scriptsize
\textbf{Takeaway:} Align regions to NAPOT boundaries to minimize padding and preserve PMP entries.
\end{figure}
    }
\end{frame}

@@

\begin{frame}{Deployment Takeaways}
\begin{block}{Best Practices}
\begin{itemize}
    \item Use NAPOT-aligned, contiguous memory regions.
    \item Consolidate workloads into fewer enclaves.
    \item Batch calls to reduce context switches.
    \item Adapt enclave usage based on task complexity.
\end{itemize}
\end{block}

\begin{alertblock}{Key Principle}
    Balance performance, isolation, and PMP resource usage for optimal efficiency.
\end{alertblock}
\end{frame}

@@

% \begin{frame}{Workload-Aware Enclave Design}
% \begin{block}{Hybrid Cryptographic Workloads}
% \begin{itemize}
%     \item Keep keys inside enclaves.
%     \item Offload bulk crypto operations outside to reduce overhead.
% \end{itemize}
% \end{block}

% \begin{block}{Target Use Cases}
% \begin{itemize}
%     \item Isolate sensitive logic (e.g., ML inference, attestation).
%     \item Avoid enclaves for high-memory or parallel tasks. (e.g., bulk cryptographic ops.)
% \end{itemize}
% \end{block}
% \end{frame}

% \begin{frame}{Enclave Concurrency and Multiplexing}
%     \uncover<1->{
% \begin{block}{Problem}
% Too many active enclaves → PMP exhaustion, context switch cost.
% \end{block}
%     }
%     \uncover<2->{
% \begin{alertblock}{Solution}
% Use enclave multiplexing: combine tasks into fewer enclaves.
% \end{alertblock}
% \begin{block}{Bonus}
% Dynamic PMP reallocation can further improve scalability.
% \end{block}
%     }
% \end{frame}

@@

% Future Work and Improvements
\begin{frame}{Hardware Outlook \& Future Directions}
    \only<1>{
\begin{block}{Limitations}
\begin{itemize}
    \item Few PMP entries.
    \item Rigid NAPOT region encoding.
\end{itemize}
\end{block}
    }
    \only<2>{
\begin{alertblock}{Future Opportunities}
\begin{itemize}
    \item More flexible PMP schemes.
    \item Crypto accelerators for enclave workloads.
    \item Software-defined memory isolation models.
\end{itemize}
\end{alertblock}
    }
\end{frame}
